\section{Introduction}
\label{sec.intro}

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language for embedded soft real-time systems that aims to offer a
concurrent, safe, and expressive alternative to C with the characteristics that
follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
Synchronous languages offer a simple run-to-completion execution model that
enables deterministic execution and make formal reasoning tractable.
For this reason, it has been successfully adopted in safety-critical real-time
embedded systems~\cite{rp.twelve}.

Previous work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\%)~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.
%, and as an
%alternative language in an undergraduate-level course on embedded systems for
%the past 6 years.

\CEU inherits the synchronous and imperative mindset of Esterel but adopts a
simpler semantics with fine-grained execution control.
%
The list that follows summarizes the semantic peculiarities of
\CEU~\cite{ceu.tecs17}:
%
\begin{itemize}
    \item Fine-grained, intra-reaction deterministic execution, which makes
          \CEU fully deterministic.
    \item Stack-based execution for internal events, which provides a limited
          but memory-bounded form of subroutines.
    \item Finalization mechanism for lines of execution, which makes abortion
          safe with regard to external resources.
\end{itemize}

In this work, we present a formal small-step structural operational
semantics for \CEU and prove that it enforces memory-bounded, deterministic,
and terminating reactions to the environment, i.e., that for a given
arbitrary timeline of input events, multiple executions of the same program
always react in bounded time and arrive at the same final finite memory
state.
%
Conceiving a formal semantics for \CEU leads to a number of capabilities and
outcomes as follows:

\begin{enumerate}
\item
    Understanding, finding corner cases, and setting compromises.
    After the semantics was complete and discussed in extent, we found a bug in
    the order of execution of statements in a specific case.
\item
    Explaining core aspects of the language in a reduced, precise, and
    unambiguous way.
    This is particularly important when comparing languages that are similar in
    the surface (e.g., \CEU and Esterel).
\item
    Implementing the language.
    Small-step operational semantics typically describe abstract machines that
    are close to concrete implementations.
    The current implementation of the \CEU scheduler is based on the formal
    semantics.
\item
    Proving properties for particular or all programs in the language.
    We prove that all programs in \CEU are memory bounded, deterministic, and
    react in finite time.
\end{enumerate}

The last item is particularly important in the context of embedded systems:

\begin{description}
\item[Memory Boundedness:]
At compile time, we can ensure that the program fits in the device's restricted
memory and that it will not grow unexpectedly during runtime.
\item[Deterministic Execution:]
We can simulate an entire program execution providing an input history with the
guarantee that it will always have the same behavior.
This can be done in negligible time in a controlled development environment by
multiple developers before deploying the application in the actual device.
\item[Terminating Reactions:]
Real-time applications must guarantee responses within specified deadlines.
A terminating semantics enforces upper bounds for all reactions and guarantees
that programs always progress with the time.
In addition, it also ensures that programs always reach an idle state amenable
to entering in standby mode, which can improve the autonomy of battery-powered
devices.
\end{description}
